Something about Performance, 2D Component of NEMO. (TBC)

### 2016-01-11 nemolite2d_perf.tex
nemolite2d paper: correct \psykal definition and start work on Conclusions

\subsection{The \psykal Approach}
Proliferation of lightweight cores is resulting in increased
complexity in the programming models required to achieve performance.
Combine with the complexity of e.g. a model simulating circulation in
the global ocean. Significant challenge in software engineering.
Propose an approach for meeting this challenge by separating the
computational science demands from the {\it e.g.} oceanographic
demands.

\begin{figure}
\centering
\caption{The \psykal Approach.}
\end{figure}

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal version of nemolite2d retains the Input/Output
capability of the original. This aids in limiting the optimisations
that can be performed on the \psykal version to those that should also
be applicable to full oceanographic models. Note that although we
retain the I/O functionality, all of the results presented in this work
carefully exclude the effects of I/O since it is compute performance
that interests us here.

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal version of nemolite2d, we made the
kernels as fine-grained as possible. This resulted in eight distinct
kernels, each of which updates a single field at a single point (since
we have chosen to use point-wise kernels). With a little bit of
tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single invoke
gives us a single PSy-layer routine which consists of applying each of
the kernels to all of the points requiring an update on the model
mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
routine then contains a doubly-nested loop around each kernel call, as
indicated by the pseudo-code in
Figure~\ref{FIG_psykal_nemolite2d_structure}.

\begin{figure}
\centering
%\includegraphics[width=85mm]{psykal_nemolite2d}
\caption{A schematic of the vanilla \psykal version of the nemolite2d code.}
\label{FIG_psykal_nemolite2d_structure}
\end{figure}

As with any full oceanographic model, boundary conditions must be
applied at the edges of the model domain. Since nemolite2d applies
external boundary conditions (e.g. barotropic forcing), this is done
via user-supplied kernels.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methodology(?)}
\label{sec_methodology}

Our aim in this work is to achieve portable performance. Consequently,
we have performed tests with the hardware and compilers listed in
Table~\ref{TABLE_compilers}. Where a compiler is available on a given
piece of hardware, the version number used in this work is specified.
The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB.

@@

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The matrix of compilers and hardware used in this work. The
  use of a compiler on a given system is indicated by the specification
  of the version of the compiler in the relevent element. No entry
  indicates that a compiler was not available/used on that system.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|c|}
\hline
                 & \multicolumn{5}{c|}{Compiler}            \\
\hline
                 & Gnu   & Intel       & Cray    & IBM    &  PGI \\
\hline
Intel Haswell    & 4.8.2 & 14.0.0      &         &        &  \\
Intel Ivy Bridge & 4.9.1 & 15.0.0.090  & 8.3.3   &        &  \\
IBM Power 8      &       &             &         &        &  \\
NVIDIA Tesla K40 &       &             &         &        &  {\bf Jeremy} \\
\hline
\end{tabular}
\end{table}

In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
used with each compiler. {\bf Explanation of the compiler flags?}
%These flags are particularly important for
%the \psykal versions of the code. The performance of the original
%version of the code is much less sensitive to the compiler
%options. This is because it consists of a single source file
%containing effectively a single routine (excluding those for I/O).

@@

Before applying any code transformations, we first benchmark the
original, serial version of the code. We also benchmark the vanilla,
unoptimised version after it has been re-structured following the
\psykal approach. In addition to this benchmarking, we profile these
versions of the code at the algorithm level (using a high-level timing
API). The resulting profiles are given in
Table~\ref{TABLE_profile}. The Momentum section dominates the profile
of both versions of the code, accounting for around 70--80\% of the
wall-clock time spent doing time-stepping. It is also the key section
when we consider the performance loss when moving from the original to
the \psykal version of the code; it slows down by a factor of
two. Although less significant in terms of absolute time, the
Continuity section is also dramatically slower in the \psykal version,
this time by a factor of three. In contrast, the performance of the
Time-update and Boundary Condition regions are not significantly
affected by the move to the \psykal version.

\begin{table}[!t]
\caption{The performance profile of the original and \psykal versions
  of nemolite2d (for 2000 time-steps of the $128^{2}$ domain and the
  Intel compiler).}
\label{TABLE_profile}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla \psykal} &
\multicolumn{2}{c}{Final \psykal} \\
        & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
\hline
Momentum    & 1.98  & 72.6  & 4.05  & 79.5 & 2.09 & 75.3 \\
Time-update & 0.40  & 14.6  & 0.41  & 8.1  & 0.29 & 10.6 \\
BCs         & 0.25  & 9.1   & 0.29  & 5.7  & 0.27 & 9.9  \\
Continuity  & 0.10  & 3.7   & 0.33  & 6.6  & 0.11 & 4.1  \\
\hline
\end{tabular}
\end{table}

Beginning with the vanilla \psykal version, we then apply a series of
code transformations while obeying the \psykal separation of concerns,
{\it i.e.} optimisation is restricted to the middle, {PS}y layer and
leaves the kernel and algorithm layers unchanged. The aim of these
optimisations is to recover, as much as is possible, the performance
of the original version of the code. The transformations we have
performed and the reasons for them are described in the following sections.

\subsubsection{Addition of `safe-address' directives}
\label{sec_safe_address}

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the nemolite2d kernels contain conditional statements. These
statements are there to check whether {\it e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
ftn-6375 ftn: VECTOR File = nemolite2d.f90,
 Line = 448 
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the field
module to allocate all field-data arrays with extents greater than
strictly required.  We were then able to safely add the safe\_address
before all of the loops where the Cray compiler indicated it might be
useful (the Momentum loops and some of the BC loops).

@@

The profiling data in Table~\ref{TABLE_profile} shows that it is the
Momentum section that accounts for the bulk of the model run-time.  We
therefore chose to attempt to optimise this section first. In-keeping
with the \psykal approach, we are only permitted to optimise the middle
(PSy) layer which, for this section comprises calls to two kernels,
one for each of the $x$ and $y$ components of momentum. These kernels
are relatively large; each comprises roughly 85 lines of Fortran
(excluding variable declarations).

@@

It turns out that the Cray-compiled binaries of both the original and
\psykal versions of nemolite2d perform considerably less well than
their Intel-compiled counterparts. Comparison of the diagnostic output
from each of the compilers revealed that while the Intel compiler was
happy to vectorise the Momentum loops, the Cray compiler was choosing
not to:
\begin{verbatim}
   99.          do ji = 2, M-1, 1
 A loop starting at line 99 was blocked 
 with block size 256.

 A loop starting at line 99 was not
 vectorized because it contains 
 conditional code which is more
 efficient if executed in scalar mode.
\end{verbatim}
Inserting the Cray `vector always' directive persuaded the compiler
to vectorise the loop:
\begin{verbatim}
  99.    !dir$ vector always
  100.          do ji = 2, M-1, 1
 A loop starting at line 100 was blocked
 with block size 256.

@@

Having optimised the Momentum section as much as permitted by the
\psykal approach, we turn our attention to the three remaining
sections of the code. The profile data in Table~\ref{TABLE_profile}
shows that these regions are all comparable in terms of cost. What is
striking however, is that the cost of the Continuity section increases
by more than a factor of three in moving to the \psykal version of the
code.

Comparison of the diagnostic output from the Cray and Intel
compilers revealed that the Cray compiler was vectorising the
Continuity section while the Intel compiler reported that it was
unable to do so due to dependencies. After some experimentation we
found that this was due to limitations in the compiler's analysis of
the way components of Fortran derived types were being used. Each
GOcean field object, in addition to the array holding the local
section of the field, contains a pointer to a GOcean grid object. If a
kernel requires grid-related quantities ({\it e.g.} the grid spacing)
then these are obtained by passing it a reference to the appropriate
array within the grid object. Although these grid-related quantities
are read-only within a compute kernel, if they were obtained from the
same object as that containing the array to which that kernel writes
then the Intel compiler identified a dependency preventing
vectorisation.  This limitation was simply removed by ensuring that
all read-only quantities were accessed via field objects that were
themselves read-only for the kernel at hand.  For instance, the call
to the continuity kernel, which confused the Intel compiler, originally
looked like this:
\begin{verbatim}
do jj = 2, N, 1
 do ji = 2, M, 1

@@

For this part of the work we began with the optimal \psykal version of
the code, as obtained after applying the various transformations
described in the previous section. As with the transformations of the
serial code, our purpose here is to determine the functionality
required of the PSyclone code-generation system.

@@

Moving now the to the \psykal version of nemolite2d,
Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
fastest \psykal version for each of the compiler/CPU combinations.
While the Intel compiler still produces the best-performing binary,
the Cray-compiled binary is now a very close second. In fact, the
performance of both the Gnu- and Cray-compiled \psykal versions is
generally significantly greater than that of their respective original
versions. We also note that best absolute performance (in terms of
grid points processed per second) with any compiler is obtained with
the $256^2$ domain.

\begin{figure}[!t]
\centering
\includegraphics[width=7.5cm]{best_psykal_summary}
\caption{Summary of the best performance achieved by any \psykal 
version of nemolite2d with each of the compilers and CPUs under 
consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

Figure~\ref{FIG_slowdown_summary} plots the percentage difference
between the performance of the original and the best \psykal versions
of nemolite2d for each compiler/CPU combination. This shows that it is
only the Intel-compiled binary running the $64^2$ domain that is
slower with the \psykal version of the code (and then only by some
3\%). For all other points in the space, the optimised \psykal version
of the code performs better. The results for the Cray compiler are
however somewhat sqewed by the fact that it did not SIMD vectorise key
parts of the original version (see below).

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{slowdown_summary}
\caption{Comparison of the performance of the best \psykal
version with that of the original version of the code. A negative value 
indicates that the \psykal version is slower than the original.}
\label{FIG_slowdown_summary}
\end{figure}

Having show that we can recover, and often improve upon, the
performance of the original version of nemolite2d, the next logical
step is to examine the necessary code transformations in detail.  We
do this for the $256^{2}$ case since this fits within cache on the Ivy
Bridge CPUs we are using here.  Table~\ref{TABLE_opt_breakdown} shows
detailed performance figures for this case after each transformation has been
applied to the code. The same data is visualised in
Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal
version for a given compiler/CPU is given relative to the performance
of the original with the same compiler/CPU.

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Performance (millions of points updated per second) for the
  $256^2$ case after each code transformation. Where an optimisation
  utilises compiler-specific directives then performance figures for
  the other compilers are omitted.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c}
\hline
Compiler:           & Gnu & Intel & Cray  \\
\hline
CPU:                & \multicolumn{3}{c}{Ivy Bridge}  \\
\hline
Original         & 3.87 & 12.2 & 6.83 \\
Vanilla \psykal  & 3.07 & 6.55 & 6.73 \\
Safe-address     & --   & --   & 6.95 \\
In-line Momentum & 4.31 & 10.7 & 6.93 \\
SIMD Momentum    & --   & --   & 11.8 \\
\parbox{2.5cm}{\raggedright Grid data from read-only objects} & 4.31 & 11.3 & 11.8 \\
In-line Continuity         & 4.83 & 11.8 & 11.6 \\
In-line remaining kernels  & 5.89 & 12.0 & 11.5 \\
In-line field copies       & 5.92 & 12.5 & 11.4 \\
\hline
\parbox{2.5cm}{\raggedright \%-speed-up of best {\it c.f.} original} & 34.62 & 2.39 & 42.22 \\
\hline
\end{tabular}
\end{table}

\begin{figure}[!t]
\centering
\includegraphics[width=90mm]{opt_stages_256}
\caption{Performance of the \psykal version of nemolite2d for the
  $256^{2}$ domain at each stage of optimisation. The first (black)
  bar of each cluster gives the performance of the original version of
  nemolite2d for that compiler/CPU combination.}
\label{FIG_opt_stages_256}
\end{figure}

@@

We now turn to transformations related to parallelisation of the
nemolite2d code; the introduction of OpenMP and OpenACC directives. In
keeping with the \psykal approach we do not modify either the
Algorithm- or Kernel-layer code.  Any code changes are restricted to
either the PSy (middle) layer or the underlying library that manages
e.g. the construction of field objects.

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_problem_size}
\caption{The scaling behaviour of the most performant OpenMP-parallel
  version of \psykal nemolite2d for the full range of domain sizes
  considered on the CPU. Results are for the Intel compiler on a
  single Intel Ivy Bridge socket. The corresponding parallel
  efficiencies are shown using open symbols and dashed lines. The
  24-thread runs employed hyperthreading.}
\label{FIG_omp_domain_size}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_gnu}
\caption{Performance of the OpenMP-parallel version of \psykal
  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
  socket. The corresponding parallel efficiencies are shown using open
  symbols and dashed lines. The 24-thread runs employed hyperthreading
  and the optimal OpenMP schedule is given in parentheses.}
\label{FIG_gnu_omp_256}
\end{figure}

@@

In restricting ourselves to a single socket, we are keeping all
threads within a single NUMA region. It is therefore surprising that
implementing a `first-touch' policy has any effect and yet, for the
Gnu- and Intel-compiled binaries, it appears to improve performance
when hyperthreading is employed to run on 24 threads (green lines and
diamond symbols in Figures~\ref{FIG_gnu_omp_256} and
~\ref{FIG_intel_omp_256}). {\bf Can we explain this?}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_intel}
\caption{Performance of the OpenMP-parallel version of \psykal
  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
  socket.  The corresponding parallel efficiencies are shown using
  open symbols and dashed lines. The 24-thread runs employed
  hyperthreading and the optimal OpenMP schedule is given in
  parentheses.}
\label{FIG_intel_omp_256}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_cray}
\caption{Performance of the OpenMP-parallel version of \psykal
  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
  socket.  The corresponding parallel efficiencies are shown using
  open symbols and dashed lines. The 24-thread runs employed
  hyperthreading and the optimal OpenMP schedule is given in
  parentheses.}
\label{FIG_cray_omp_256}
\end{figure}

@@

Since NEMO and similar finite-difference codes tend to be
memory-bandwidth bound, we checked the sensitivity of our performance
results to this quantity by benchmarking using two sockets of Intel
Ivy Bridge ({\it i.e.} using a complete node of Archer). For this
configuration we ensured that threads were evenly shared over the two
sockets. The performance obtained for the $256^2$ case with the `early
SINGLE' version of the code is compared with the single-socket
performance in Figure~\ref{FIG_omp_2socks}. Surprisingly, doubling the
available memory bandwidth in this way has little effect on
performance - the two-socket performance figures track those from a
single socket very closely. The only realy difference in performance
is at 24 threads where, in addition to the difference in available
memory bandwidth, the single-socket configuration is using
hyperthreading while the two-socket case is not.  {\bf Have query in
  with Archer helpdesk on behaviour seen with Cray compiler here. Also
  would like to run 512 case on 2 sockets to see how it goes out to 48
  threads.}

>>>>

\subsection{The \psykal\ Approach}
Proliferation of lightweight cores is resulting in increased
complexity in the programming models required to achieve performance.
Combine with the complexity of e.g. a model simulating circulation in
the global ocean. Significant challenge in software engineering.
Propose an approach for meeting this challenge by separating the
computational science demands from the {\it e.g.} oceanographic
demands.

\begin{figure}
\centering
\caption{The \psykal\ Approach.}
\end{figure}

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal\ version of nemolite2d retains the Input/Output
capability of the original. This aids in limiting the optimisations
that can be performed on the \psykal\ version to those that should also
be applicable to full oceanographic models. Note that although we
retain the I/O functionality, all of the results presented in this work
carefully exclude the effects of I/O since it is compute performance
that interests us here.

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal\ version of nemolite2d, we made the
kernels as fine-grained as possible. This resulted in eight distinct
kernels, each of which updates a single field at a single point (since
we have chosen to use point-wise kernels). With a little bit of
tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single invoke
gives us a single PSy-layer routine which consists of applying each of
the kernels to all of the points requiring an update on the model
mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
routine then contains a doubly-nested loop around each kernel call, as
indicated by the pseudo-code in
Figure~\ref{FIG_psykal_nemolite2d_structure}.

\begin{figure}
\centering
%\includegraphics[width=85mm]{psykal_nemolite2d}
\caption{A schematic of the vanilla \psykal\ version of the nemolite2d code.}
\label{FIG_psykal_nemolite2d_structure}
\end{figure}

As with any full oceanographic model, boundary conditions must be
applied at the edges of the model domain. Since nemolite2d applies
external boundary conditions ({\it e.g.} barotropic forcing), this is
done via user-supplied kernels.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methodology}
\label{sec_methodology}

Our aim in this work is to achieve portable performance. Consequently,
we have performed tests with the hardware and compilers listed in
Table~\ref{TABLE_compilers}. Where a compiler is available on a given
piece of hardware, the version number used in this work is specified.
%The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
%3.7~GHz and 10~MB of last-level cache. 
The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB.

@@

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
%
\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The matrix of compilers and hardware used in this work. The
  use of a compiler on a given system is indicated by the specification
  of the version of the compiler in the relevent element. No entry
  indicates that a compiler was not available/used on that system.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|c|}
\hline
                 & \multicolumn{5}{c|}{Compiler}            \\
\hline
                 & Gnu   & Intel       & Cray    & IBM    &  PGI \\
\hline
%Intel Haswell    & 4.8.2 & 14.0.0      &         &        &  \\
Intel Ivy Bridge & 4.9.1 & 15.0.0.090  & 8.3.3   &        &  \\
IBM Power 8      &       &             &         &        &  \\
NVIDIA Tesla K40 &       &             &         &        &  {\bf Jeremy} \\
\hline
\end{tabular}
\end{table}

In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
used with each compiler. {\bf Explanation of the compiler flags?}
%These flags are particularly important for
%the \psykal\ versions of the code. The performance of the original
%version of the code is much less sensitive to the compiler
%options. This is because it consists of a single source file
%containing effectively a single routine (excluding those for I/O).

@@

Before applying any code transformations, we first benchmark the
original, serial version of the code. We also benchmark the vanilla,
unoptimised version after it has been re-structured following the
\psykal\ approach. In addition to this benchmarking, we profile these
versions of the code at the algorithm level (using a high-level timing
API). The resulting profiles are given in
Table~\ref{TABLE_profile}. The Momentum section dominates the profile
of both versions of the code, accounting for around 70--80\% of the
wall-clock time spent doing time-stepping. It is also the key section
when we consider the performance loss when moving from the original to
the \psykal\ version of the code; it slows down by a factor of
two. Although less significant in terms of absolute time, the
Continuity section is also dramatically slower in the \psykal\ version,
this time by a factor of three. In contrast, the performance of the
Time-update and Boundary Condition regions are not significantly
affected by the move to the \psykal\ version.

\begin{table}[!t]
\caption{The performance profile of the original and \psykal\ versions
  of nemolite2d on the Intel Ivy Bridge CPU (for 2000 time-steps of the
  $128^{2}$ domain and the Intel compiler).}
\label{TABLE_profile}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla \psykal} &
\multicolumn{2}{c}{Final \psykal} \\
        & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
\hline
Momentum    & 1.98  & 72.6  & 4.05  & 79.5 & 2.09 & 75.3 \\
Time-update & 0.40  & 14.6  & 0.41  & 8.1  & 0.29 & 10.6 \\
BCs         & 0.25  & 9.1   & 0.29  & 5.7  & 0.27 & 9.9  \\
Continuity  & 0.10  & 3.7   & 0.33  & 6.6  & 0.11 & 4.1  \\
\hline
\end{tabular}
\end{table}

Beginning with the vanilla \psykal\ version, we then apply a series of
code transformations while obeying the \psykal\ separation of concerns,
{\it i.e.} optimisation is restricted to the middle, {PS}y layer and
leaves the kernel and algorithm layers unchanged. The aim of these
optimisations is to recover, as much as is possible, the performance
of the original version of the code. The transformations we have
performed and the reasons for them are described in the following sections.

\subsubsection{Addition of {\tt safe\_address} directives}
\label{sec_safe_address}

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the nemolite2d kernels contain conditional statements. These
statements are there to check whether {\it e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
ftn-6375 ftn: VECTOR File = nemolite2d.f90,
 Line = 448 
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the field
module to allocate all field-data arrays with extents greater than
strictly required.  We were then able to safely add the {\tt safe\_address}
before all of the loops where the Cray compiler indicated it might be
useful (the Momentum loops and some of the BC loops).

@@

The profiling data in Table~\ref{TABLE_profile} shows that it is the
Momentum section that accounts for the bulk of the model run-time.  We
therefore chose to attempt to optimise this section first. In-keeping
with the \psykal\ approach, we are only permitted to optimise the middle
(PSy) layer which, for this section comprises calls to two kernels,
one for each of the $x$ and $y$ components of momentum. These kernels
are relatively large; each comprises roughly 85 lines of Fortran
(excluding variable declarations).

@@

It turns out that the Cray-compiled binaries of both the original and
\psykal\ versions of nemolite2d perform considerably less well than
their Intel-compiled counterparts. Comparison of the diagnostic output
from each of the compilers revealed that while the Intel compiler was
happy to vectorise the Momentum loops, the Cray compiler was choosing
not to:
\begin{verbatim}
   99.          do ji = 2, M-1, 1
 A loop starting at line 99 was blocked 
 with block size 256.

 A loop starting at line 99 was not
 vectorized because it contains 
 conditional code which is more
 efficient if executed in scalar mode.
\end{verbatim}
Inserting the Cray {\tt vector always} directive persuaded the compiler
to vectorise the loop:
\begin{verbatim}
  99.    !dir$ vector always
  100.          do ji = 2, M-1, 1
 A loop starting at line 100 was blocked
 with block size 256.

@@

Having optimised the Momentum section as much as permitted by the
\psykal\ approach, we turn our attention to the three remaining
sections of the code. The profile data in Table~\ref{TABLE_profile}
shows that these regions are all comparable in terms of cost. What is
striking however, is that the cost of the Continuity section increases
by more than a factor of three in moving to the \psykal\ version of the
code.

Comparison of the diagnostic output from the Cray and Intel
compilers revealed that the Cray compiler was vectorising the
Continuity section while the Intel compiler reported that it was
unable to do so due to dependencies. After some experimentation we
found that this was due to limitations in the compiler's analysis of
the way components of Fortran derived types were being used. Each
GOcean field object, in addition to the array holding the local
section of the field, contains a pointer to a GOcean grid object. If a
kernel requires grid-related quantities ({\it e.g.} the grid spacing)
then these are obtained by passing it a reference to the appropriate
array within the grid object. Although these grid-related quantities
are read-only within a compute kernel, if they were referenced from the
same field object as that containing an array to which the kernel writes
then the Intel compiler identified a dependency preventing
vectorisation.  This limitation was simply removed by ensuring that
all read-only quantities were accessed via field objects that were
themselves read-only for the kernel at hand.  For instance, the call
to the continuity kernel, which confused the Intel compiler, originally
looked like this:
\begin{verbatim}
do jj = 2, N, 1
 do ji = 2, M, 1

@@

For this part of the work we began with the optimal \psykal\ version of
the code, as obtained after applying the various transformations
described in the previous section. As with the transformations of the
serial code, our purpose here is to determine the functionality
required of the PSyclone code-generation system.

@@

Moving now the to the \psykal\ version of nemolite2d,
Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
fastest \psykal\ version for each of the compiler/CPU combinations.
While the Intel compiler still produces the best-performing binary,
the Cray-compiled binary is now a very close second. In fact, the
performance of both the Gnu- and Cray-compiled \psykal\ versions is
generally significantly greater than that of their respective original
versions. We also note that best absolute performance (in terms of
grid points processed per second) with any compiler is obtained with
the $256^2$ domain.

\begin{figure}[!t]
\centering
\includegraphics[width=7.5cm]{best_psykal_summary}
\caption{Summary of the best performance achieved by any \psykal\ 
version of nemolite2d with each of the compilers and CPUs under 
consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

Figure~\ref{FIG_slowdown_summary} plots the percentage difference
between the performance of the original and the best \psykal\ versions
of nemolite2d for each compiler/CPU combination. This shows that it is
only the Intel-compiled binary running the $64^2$ domain that is
slower with the \psykal\ version of the code (and then only by some
3\%). For all other points in the space, the optimised \psykal\ version
of the code performs better. The results for the Cray compiler are
however somewhat sqewed by the fact that it did not SIMD vectorise key
parts of the original version (see below).

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{slowdown_summary}
\caption{Comparison of the performance of the best \psykal\
version with that of the original version of the code. A negative value 
indicates that the \psykal\ version is slower than the original.}
\label{FIG_slowdown_summary}
\end{figure}

Having show that we can recover, and often improve upon, the
performance of the original version of nemolite2d, the next logical
step is to examine the necessary code transformations in detail.  We
do this for the $256^{2}$ case since this fits within cache on the Ivy
Bridge CPUs we are using here.  Table~\ref{TABLE_opt_breakdown} shows
detailed performance figures for this case after each transformation has been
applied to the code. The same data is visualised in
Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal\
version for a given compiler/CPU is given relative to the performance
of the original with the same compiler/CPU.

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Performance (millions of points updated per second) for the
  $256^2$ case after each code transformation. Where an optimisation
  utilises compiler-specific directives then performance figures for
  the other compilers are omitted.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c}
\hline
Compiler:           & Gnu & Intel & Cray  \\
\hline
CPU:                & \multicolumn{3}{c}{Ivy Bridge}  \\
\hline
Original         & 3.87 & 12.2 & 6.83 \\
Vanilla \psykal  & 3.07 & 6.55 & 6.73 \\
Safe-address     & --   & --   & 6.95 \\
In-line Momentum & 4.31 & 10.7 & 6.93 \\
SIMD Momentum    & --   & --   & 11.8 \\
\parbox{2.5cm}{\raggedright Grid data from read-only objects} & 4.31 & 11.3 & 11.8 \\
In-line Continuity         & 4.83 & 11.8 & 11.6 \\
In-line remaining kernels  & 5.89 & 12.0 & 11.5 \\
In-line field copies       & 5.92 & 12.5 & 11.4 \\
\hline
\parbox{2.5cm}{\raggedright \%-speed-up of best {\it c.f.} original} & 34.6 & 2.39 & 42.2 \\
\hline
\end{tabular}
\end{table}

\begin{figure}[!t]
\centering
\includegraphics[width=93mm]{opt_stages_256}
\caption{Performance of the \psykal\ version of nemolite2d for the
  $256^{2}$ domain at each stage of optimisation. The first (black)
  bar of each cluster gives the performance of the original version of
  nemolite2d for that compiler/CPU combination.}
\label{FIG_opt_stages_256}
\end{figure}

@@

We now turn to transformations related to parallelisation of the
nemolite2d code; the introduction of OpenMP and OpenACC directives. In
keeping with the \psykal\ approach we do not modify either the
Algorithm- or Kernel-layer code.  Any code changes are restricted to
either the PSy (middle) layer or the underlying library that manages
{\it e.g.} the construction of field objects.

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_problem_size}
\caption{The scaling behaviour of the most performant OpenMP-parallel
  version of \psykal\ nemolite2d for the full range of domain sizes
  considered on the CPU. Results are for the Intel compiler on a
  single Intel Ivy Bridge socket. The corresponding parallel
  efficiencies are shown using open symbols and dashed lines. The
  24-thread runs employed hyperthreading.}
\label{FIG_omp_domain_size}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_gnu}
\caption{Performance of the OpenMP-parallel version of \psykal\
  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
  socket. The corresponding parallel efficiencies are shown using open
  symbols and dashed lines. The 24-thread runs employed hyperthreading
  and the optimal OpenMP schedule is given in parentheses.}
\label{FIG_gnu_omp_256}
\end{figure}

@@

In restricting ourselves to a single socket, we are keeping all
threads within a single NUMA region. It is therefore surprising that
implementing a `first-touch' policy has any effect and yet, for the
Gnu- and Intel-compiled binaries, it appears to improve performance
when hyperthreading is employed to run on 24 threads (green lines and
diamond symbols in Figures~\ref{FIG_gnu_omp_256} and
~\ref{FIG_intel_omp_256}). {\bf Can we explain this? Perhaps with HW
  counters?}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_intel}
\caption{Performance of the OpenMP-parallel version of \psykal\
  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
  socket.  The corresponding parallel efficiencies are shown using
  open symbols and dashed lines. The 24-thread runs employed
  hyperthreading and the optimal OpenMP schedule is given in
  parentheses.}
\label{FIG_intel_omp_256}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_cray}
\caption{Performance of the OpenMP-parallel version of \psykal\
  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
  socket.  The corresponding parallel efficiencies are shown using
  open symbols and dashed lines. The 24-thread runs employed
  hyperthreading and the optimal OpenMP schedule is given in
  parentheses.}
\label{FIG_cray_omp_256}
\end{figure}

@@

Since NEMO and similar finite-difference codes tend to be
memory-bandwidth bound, we checked the sensitivity of our performance
results to this quantity by benchmarking using two sockets of Intel
Ivy Bridge ({\it i.e.} using a complete node of Archer). For this
configuration we ensured that threads were evenly shared over the two
sockets. The performance obtained for the $256^2$ case with the `early
SINGLE' version of the code is compared with the single-socket
performance in Figure~\ref{FIG_omp_2socks}. Surprisingly, doubling the
available memory bandwidth in this way has little effect on
performance - the two-socket performance figures track those from a
single socket very closely. The only significant difference in performance
is at 24 threads where, in addition to the difference in available
memory bandwidth, the single-socket configuration is using
hyperthreading while the two-socket case is not.  {\bf Have query in
  with Archer helpdesk on behaviour seen with Cray compiler here. Also
  would like to run 512 case on 2 sockets to see how it goes out to 48
  threads.}

{\bf Add roofline model to aid understanding of this result.}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{1and2sockets}
\caption{Performance of the OpenMP-parallel version of \psykal\
  nemolite2d on one and two sockets of Intel Ivy Bridge. The 24-thread
  runs on a single socket used hyperthreading and the two-socket runs
  had the threads shared equally between the sockets.}
\label{FIG_omp_2socks}
\end{figure}
